\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-16 jasa/paper-jasa.tex
text for diagnostic plots

While Theorem~\ref{thm:ratio-equivalence} states that the likelihood ratio $r(\mathbf{x}; \theta_0, \theta_1)$ is
invariant under the dimensionality reduction $s(\mathbf{x}; \theta_0, \theta_1)$ provided that it is monotonic with
$r(\mathbf{x}; \theta_0, \theta_1)$ itself and we know that any universally strongly consistent algorithm can be
used to learn such a function, we know that in practice $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ will not be exact.
Thus, it is crucial that to have some diagnostic procedures to assess the quality of this approximation.
This is complicated by the fact that in the likelihood-free setting, we don't have access to the true likelihood ratio.
Below we consider two such diagnostic procedures that can be implemented in the likelihood-free setting.

The first diagnostic procedure is related to the procedure for finding the MLE $\hat{\theta}$ in Eqn.~\ref{eq:mle_withs}. As pointed out there it is important that one maximizes the likelihood ratio as the surface integral and Jacobian factors related to the dimensionality reduction only cancel in the ratio (see Eqn.~\ref{eqn:factorization}). Importantly, they also only cancel if the dimensionality reduction map satisfies the conditions needed for Theorem~\ref{thm:ratio-equivalence} to hold. Moreover, the resulting value of $\hat{\theta}$ should be independent of the value of $\theta_1$ used in the denominator of the likelihood ratio. Similarly, we have
%$\log\Lambda(\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\hat\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\theta_1) -\log p(\mathcal{D}|\hat\theta)/p(\mathcal{D}|\theta_1)$ for all values of $\theta_1$.
\begin{equation}
\log\Lambda(\theta) = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\hat\theta)} = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\theta_1)} -\log \frac{p(\mathcal{D}|\hat\theta)}{p(\mathcal{D}|\theta_1)}
\end{equation}
for all values of $\theta_1$. Thus, by explicitly checking the independence of these quantities on $\theta_1$ we indirectly probe the quality of the approximation $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1)) \approx {r}({s}(\mathbf{x}; \theta_0, \theta_1))$.

@@

We illustrate these diagnostic procedures in Fig.~\ref{fig:ratio_comp}.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
        \label{fig:1a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
        \label{fig:p_sij}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using random forest}
        \label{fig:1d}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1e.pdf}
        \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
        \label{fig:decomposed_ratio}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1f.pdf}
        \caption{$-2\log \Lambda(\gamma)$ }
        \label{fig:likelihood}
    \end{subfigure}

@@

    \begin{subfigure}[t]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
        \caption{Exact vs. approximated MLEs.}
        \label{fig:2a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
        \caption{$p(-2 \log \Lambda(\gamma=0.05) \, | \, \gamma=0.05)$}
        \label{fig:2b}
    \end{subfigure}

@@

Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values. Some roughness in the contours is observed, which is primarily due to variance introduced in the calibration procedure. In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary. As discussed in Section~\ref{sec:param-calibration}, a parameterized calibration
procedure should ameliorate this issue, but that is left for now as an area for future work.
Nevertheless, optimizing the approximate log-likelihood ratio with
a Bayesian optimization~\citep{brochu2010tutorial} procedure is efficient and effective.
After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
While the goal of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.
%For comparison, Figure~\ref{fig:5b} shows the approximate log likelihood ratio evaluated on a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
%roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=10.075em]{figures/likelihood_comp_0.pdf}
        \caption{Poorly trained, well calibrated. }
        \label{fig:6a}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=9.5em]{figures/ROC_comp0.pdf}
        \caption{Poorly trained, well calibrated. }
        \label{fig:6b}
    \end{subfigure}

    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=10.075em]{figures/likelihood_comp_1.pdf}
        \caption{Poorly calibrated, well trained. }
        \label{fig:6c}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=9.5em]{figures/ROC_comp1.pdf}
        \caption{Poorly calibrated, well trained. }
        \label{fig:6d}
    \end{subfigure}

    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=10.075em]{figures/likelihood_comp_2.pdf}
        \caption{Well trained, well calibrated. }
        \label{fig:6e}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[height=9.5em]{figures/ROC_comp2.pdf}
        \caption{Well trained, well calibrated. }
        \label{fig:6f}
    \end{subfigure}

>>>>

While Theorem~\ref{thm:ratio-equivalence} states that the likelihood ratio $r(\mathbf{x}; \theta_0, \theta_1)$ is
invariant under the dimensionality reduction $s(\mathbf{x}; \theta_0, \theta_1)$ provided that it is monotonic with
$r(\mathbf{x}; \theta_0, \theta_1)$ itself and we know that any universally strongly consistent algorithm can be
used to learn such a function, we know that in practice $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ will not be exact.
Thus, it is crucial that to have some diagnostic procedures to assess the quality of this approximation.
This is complicated by the fact that in the likelihood-free setting, we don't have access to the true likelihood ratio.
Below we consider two such diagnostic procedures that can be implemented in the likelihood-free setting. We illustrate these diagnostic procedures in Fig.~\ref{fig:diagnostics}.

The first diagnostic procedure is related to the procedure for finding the MLE $\hat{\theta}$ in Eqn.~\ref{eq:mle_withs}. As pointed out there it is important that one maximizes the likelihood ratio as the surface integral and Jacobian factors related to the dimensionality reduction only cancel in the ratio (see Eqn.~\ref{eqn:factorization}). Importantly, they also only cancel if the dimensionality reduction map satisfies the conditions needed for Theorem~\ref{thm:ratio-equivalence} to hold. Moreover, the resulting value of $\hat{\theta}$ should be independent of the value of $\theta_1$ used in the denominator of the likelihood ratio. Similarly, we have
%$\log\Lambda(\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\hat\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\theta_1) -\log p(\mathcal{D}|\hat\theta)/p(\mathcal{D}|\theta_1)$ for all values of $\theta_1$.
\begin{equation}\label{eq:theta_1_independence}
\log\Lambda(\theta) = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\hat\theta)} = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\theta_1)} -\log \frac{p(\mathcal{D}|\hat\theta)}{p(\mathcal{D}|\theta_1)}
\end{equation}
for all values of $\theta_1$. Thus, by explicitly checking the independence of these quantities on $\theta_1$ we indirectly probe the quality of the approximation $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1)) \approx {r}({s}(\mathbf{x}; \theta_0, \theta_1))$.

@@

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm, width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
        \label{fig:1a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1b.pdf}
        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
        \label{fig:p_sij}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1c.pdf}
        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1d.pdf}
        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using random forest}
        \label{fig:1d}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1e.pdf}
        \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
        \label{fig:decomposed_ratio}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1f.pdf}
        \caption{$-2\log \Lambda(\gamma)$ }
        \label{fig:likelihood}
    \end{subfigure}

@@

    \begin{subfigure}[t]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2a.pdf}
        \caption{Exact vs. approximated MLEs.}
        \label{fig:2a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2b.pdf}
        \caption{$p(-2 \log \Lambda(\gamma=0.05) \, | \, \gamma=0.05)$}
        \label{fig:2b}
    \end{subfigure}

@@

Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values. Some roughness in the contours is observed, which is primarily due to variance introduced in the calibration procedure. In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary. As discussed in Section~\ref{sec:param-calibration}, a parameterized calibration
procedure should ameliorate this issue, but that is left for now as an area for future work.
Nevertheless, optimizing the approximate log-likelihood ratio with
a Bayesian optimization~\citep{brochu2010tutorial} procedure is efficient and effective.
After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
While the objective of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.
%For comparison, Figure~\ref{fig:5b} shows the approximate log likelihood ratio evaluated on a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
%roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.

Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for this example. To aid in visualization, we restrict to a 1-dimensional slice of the likelihood along $\alpha$ with $\beta=1$.   We consider three situations: i) a poorly trained, but well calibrated classifier; ii) a well trained, but poorly calibrated classifier; and iii) a well trained, and well calibrated classifier. For each case, we employ two diagnostic tests. The first checks for independence of $-2\log\Lambda(\theta)$ with respect to changes in the reference value $\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}. The second uses a classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x}; \theta_0, \theta_1)$.  As discussed for Fig.~\ref{fig:5b}, statistical fluctuations in the calibration lead to some noise in the raw approximate likelihood. Thus, we show the posterior mean of a Gaussian processes resulting from Bayesian optimization of  the raw approximate likelihood as in Fig.~\ref{fig:5c}. In addition, the standard deviation of the Gaussian process is shown for one of the $\theta_1$ reference points to indicate the size of these statistical fluctuations. It is clear that in the well calibrated cases that these fluctuations are small, while in the poorly calibrated case these fluctuations are large. Moreover, in Fig.~\ref{fig:5a} we see that in the poorly trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ has a significant dependence on the $\theta_1$ reference point. In contrast, in Fig.~\ref{fig:5c} the likelihood curves vary significantly, but this is comparable to the fluctuations expected from the calibration procedure. Finally, Fig.~\ref{fig:5e} shows that in the well trained, well calibrated case that the likelihood curves are all consistent with the exact likelihood within the estimated uncertainty band of the Gaussian process.  The ROC curves tell a similarly revealing story. As expected, the classifier is not able to distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted by the exact likelihood ratio. We can also rule out that this is a deficiency in the classifier because the two distributions are well separated when no weights are applied to $p(\mathbf{x}|\theta_1)$. In both Fig.~\ref{fig:5b} and Fig.~\ref{fig:5d} the ROC curve correctly diagnoses deficiencies in the approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$. Finally, Fig.~\ref{fig:5f} shows that the ROC curve in the well trained, well calibrated case is almost identical with the exact likelihood ratio, confirming the quality of the approximation.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_0.pdf}
        \caption{Poorly trained, well calibrated. }
        \label{fig:6a}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp0.pdf}
        \caption{Poorly trained, well calibrated. }
        \label{fig:6b}
    \end{subfigure}

    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_1.pdf}
        \caption{Poorly calibrated, well trained. }
        \label{fig:6c}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp1.pdf}
        \caption{Poorly calibrated, well trained. }
        \label{fig:6d}
    \end{subfigure}

    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_2.pdf}
        \caption{Well trained, well calibrated. }
        \label{fig:6e}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp2.pdf}
        \caption{Well trained, well calibrated. }
        \label{fig:6f}
    \end{subfigure}
